A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2022-12-28 paper/fascd.tex
done with edits section 8

In summary, inclusions \eqref{eq:fe:downwardsuminclusion} and \eqref{eq:fe:upwardsuminclusion} suffice to show the admissibility of all iterates and corrections in V-cycle Algorithm \ref{alg:nmcd}.  Lemma \ref{lem:downwardadmissibility} explains downward admissibility, and Lemma \ref{lem:upwardadmissibility} upward admissibility, in Algorithm \ref{alg:nmcd}.  However, a convergence analysis following \cite{Tai2003} or \cite{GraeserKornhuber2009}, somehow extending their arguments to VIs \eqref{eq:vi} which are not of minimization type, and/or to box constraints, would seem to require full CDs, and perhaps new ideas.

@@

To state \fascd Algorithm \ref{alg:nmcd} below we must clarify the problems solved at each mesh level.  As in FAS multigrid for nonlinear PDEs \cite{BrandtLivne2011,Bruneetal2015,Trottenbergetal2001}, both the solution approximation and the residual must be coarsened, and a source functional created.  On mesh level $j$ we assume a re-discretized nonlinear operator $f^j$, with output in $(\cV^j)'$, which is an FE discretization of $f$ in \eqref{eq:boxdirichletvi}, presumably using quadrature and etc.~in the same scheme which generated $f^J$ in \eqref{eq:fe:vi}.  Note that finest-level problem \eqref{eq:fe:vi} uses the original source functional $\ell^J$, and an admissible iterate $w^J \in \mathcal{K}^J$ has already\footnote{Conceptually-speaking this is true.  In practice the LDCs $\underline{\chi}^j,\overline{\chi}^j,\underline{\phi}^j,\overline{\phi}^j$ are generated during the descent stage; see Algorithm \ref{alg:nmcd}.  Observe that the LDCs do not depend on $w^j$ for $j<J$, nor on the corrections $y^j,z^j$.} been used to general the LDCs and constraint sets $\mathcal{D}^j$, $\mathcal{U}^j$ at each level (Section \ref{sec:ldcs}).

@@

If we were to inject the original box constraints \eqref{eq:fe:fineconstraintset} downward to coarser meshes, i.e.~$\underline{\gamma}^j = \iR \underline{\gamma}^{j+1}$ and $\overline{\gamma}^j = \iR \overline{\gamma}^{j+1}$, then, because injection preserves inequalities, and because $y^{j+1} \in \mathcal{D}^{j+1}$, it would follow from \eqref{eq:fe:definew} that $\underline{\gamma}^j \le w^j \le \overline{\gamma}^j$.  In this sense $w^j$ is admissible with respect to the box constraints, but, following \cite{GraeserKornhuber2009}, our use of LDCs (Table \ref{tab:ldcs}) avoids the need to generate such box constraints $\underline{\gamma}^j$ and $\overline{\gamma}^j$ on each level.  (Note that while $w^j \in \cV^j \subset \cV^J$ holds, generally $w^j \notin \cK^J$.)  In any case, Algorithm \ref{alg:nmcd} will yield a final, finest-level iterate $w^J+z^J \in \cK^J$ which is genuinely admissible for problem \eqref{eq:fe:vi}.

Regarding formula \eqref{eq:fe:levelsource}, the source function $\ell^j$ is determined by a classical FAS-type formula.  (Compare equation (8.5b) in \cite{BrandtLivne2011} or equation (5.3.14) in \cite{Trottenbergetal2001}.)  It has the following explanation, mimicking the PDE story from textbooks, which simultaneously justifies \eqref{eq:fe:levelsource} and VI problems \eqref{eq:fe:downvi} and \eqref{eq:fe:upvi}.  First, the $j+1$ level problem \eqref{eq:fe:downvi} is not solved exactly by the finer-level correction $y^{j+1}$; the down-smoother was only approximate.  Conceptually, we seek a further correction $y$ on the $j+1$ level so that
\begin{equation}
\ip{f^{j+1}(g^{j+1}+y^{j+1}+y)}{v-y^{j+1}-y} \ge \ip{\ell^{j+1}}{v-y^{j+1}-y}, \, \forall v\in \mathcal{D}^j \quad \text{\emph{(notional)}} \label{eq:fe:downvinotional}
\end{equation}
holds exactly.  However, $y$ will actually be approximated from a problem on the next-coarser level.  This $j$th-level problem comes from modifying notional VI \eqref{eq:fe:downvinotional} in 4 steps:
\begin{enumerate}
\item subtract the computable quantity $f^{j+1}(g^{j+1}+y^{j+1}) \in (\mathcal{V}^{j+1})'$ from both sides,
\item replace the residual $\ell^{j+1}-f^{j+1}(g^{j+1}+y^{j+1})$, now on the right, by its restriction,
\item replace $g^{j+1}+y^{j+1}$, where it appears on the left, by its restriction,
\item and, where it appears on the left, replace $f^{j+1}$ by its coarser rediscretization $f^j$.
\end{enumerate}
Steps \emph{(ii)}--\emph{(iv)} are justified, as for PDEs \cite[subsection 5.3.4, for example]{Trottenbergetal2001}, because the residual and the error should be smooth as a result of the progress made by the $j+1$ level down-smoother.  These steps now yield a problem for the $j$th-level correction $y^j \in \mathcal{D}^j$:
\begin{align}
&\ip{f^j\left(\iR(g^{j+1}+y^{j+1})+y^j\right)}{v-y^j} - \ip{f^j\left(\iR(g^{j+1}+y^{j+1})\right)}{v-y^j} \label{eq:fe:downvicluttered} \\
&\qquad \ge \ip{R\left(\ell^{j+1}-f^{j+1}(g^{j+1}+y^{j+1})\right)}{v-y^j} \notag
\end{align}
This is the same as VI problem \eqref{eq:fe:downvi}, but in cluttered form; application of definitions \eqref{eq:fe:defineg} and \eqref{eq:fe:levelsource} yields \eqref{eq:fe:downvi}.  The explanation for VI \eqref{eq:fe:upvi} is essentially identical.

We will regard the smoothers for VI problems \eqref{eq:fe:downvi} and \eqref{eq:fe:upvi} as approximate, in-place procedures, acting on the corrections $y^j$ and $z^j$, respectively.  Also, as is standard in multigrid, at the coarsest $j=0$ level we suppose that problem \eqref{eq:fe:upvi} is solved quite accurately.  However, this coarsest-level solver may be the same as the other smoothers, perhaps with a stronger tolerance.

These ideas come together in Algorithm \ref{alg:nmcd}, below.  The \pr{nmcd-vcycle} procedure acts in-place on $w^J$, but it leaves its other inputs unchanged.

\begin{pseudofloat}[ht]
\begin{pseudo} \label{ps:fascd-vcycle}
\pr{fascd-vcycle}(\ell^J,\underline{\gamma}^J,\overline{\gamma}^J;w^J)\text{:} \\+
    $\underline{\chi}^J = \underline{\gamma}^J - w^J {\large \strut}$ \\
    $\overline{\chi}^J = \overline{\gamma}^J - w^J {\large \strut}$ \\
    $g^J = w^J$ \\
    for $j=J$ downto $j=1$ \\+
      $\underline{\chi}^{j-1} = \maxR \underline{\chi}^j {\large \strut}$ \\
      $\overline{\chi}^{j-1} = \minR \overline{\chi}^j {\large \strut}$ \\
      $\underline{\phi}^j = \underline{\chi}^j - P\underline{\chi}^{j-1} {\large \strut}$ \\
      $\overline{\phi}^j = \overline{\chi}^j - P\overline{\chi}^{j-1} {\large \strut}$ \\
      $y^j = 0$ \\
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j,\underline{\phi}^j,\overline{\phi}^j,g^j;y^j)$  \ct{smoothing of \eqref{eq:fe:downvi} in $\mathcal{D}^j$}\\
      $g^{j-1} = \iR(g^j + y^j)$ \\
      $\ell^{j-1} = f^{j-1}(g^{j-1}) + R \left(\ell^j - f^j(g^j+y^j)\right)$ \\-
    $z^0 = 0$ \\
    $\text{\pr{solve}}(\ell^0,\underline{\chi}^0,\overline{\chi}^0,g^0;z^0)$ \hspace{1.0cm} \ct{solving of \eqref{eq:fe:upvi} in $\mathcal{U}^0$} \\
    for $j=1$ to $j=J$ \\+
      $z^j = y^{j} + P z^{j-1}$ \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j,\underline{\chi}^j,\overline{\chi}^j,g^j;z^j)$  \ct{smoothing of \eqref{eq:fe:upvi} in $\mathcal{U}^j$} \\-
    $w^J \gets w^J+z^J$
\end{pseudo}
\caption{The \fascd V-cycle as an iterative solver for FE VI problem \eqref{eq:fe:vi}.  $f^j$ denotes an FE discretization of $f$ in problem \eqref{eq:boxdirichletvi}.}
\label{alg:nmcd}
\end{pseudofloat}

Regarding its efficiency, to begin we consider storage.  One application of \pr{nmcd-vcycle}, assuming all box constraints are nontrivial, storage must be allocated for the seven vectors $\underline{\chi}^j$, $\overline{\chi}^j$, $\underline{\phi}^j$, $\overline{\phi}^j$, $g^j$, $\ell^j$, and $y^j$ on each level.  (Note $z^J$ can use the same storage as $y^j$.)  On the finest level one must also store $\underline{\gamma}^J$ and $\overline{\gamma}^J$, but note $g^J=w^J$ are the same.  On the coarsest level there are only five vectors to store.  Therefore, using estimates for 2D refinement plus a standard geometric series argument \cite{Trottenbergetal2001}, the total storage of a deep $V$-cycle, excluding the implementations of smoothers and the coarse-level solver, is about
\begin{equation}
9 m_J + 7 m_{J-1} + \dots + 7 m_1 + 5 m_0 \approx \left(9 + 7 \sum_{k=1}^\infty \frac{1}{4^k}\right) m_J \approx 11 m_J
\end{equation}
real numbers.  For comparison, a single-level method needs at least $4 m_J$ storage for the vectors $\underline{\gamma}^J,\overline{\gamma}^J,\ell^J,w^J$.  If some of the obstacles are trivial, for instance if $\overline{\gamma}^J=+\infty$ in a unilateral obstacle problem, storage is correspondingly reduced at all levels.

\pr{nmcd-vcycle} assumes certain semantics for the smoothers \pr{smooth} and the coarsest-level solver \pr{solve}.  These solvers do in-place modifications of their final arguments, without modifying their first four arguments.  $\pr{smooth}^{\id{down}}$ and $\pr{smooth}^{\id{up}}$ are in-exact solvers of VIs \eqref{eq:fe:downvi} and \eqref{eq:fe:upvi}, respectively, iterated \id{down} and \id{up} times.  Further implementation details are addressed in Section \ref{sec:smoothers}.

Repeated application of \pr{nmcd-vcycle} to solve VI \eqref{eq:fe:vi} will not reduce the finest-level residual $f^J(w^J) - \ell^J$ to zero everywhere, and we need to identify the vector quantity whose norm must be controlled by a tolerance.  Recalling the comments following mixed complementarity problem \eqref{eq:fe:mcp}, and denoting the hat function at node $x_p \in \mathcal{T}^J$ by $\psi_p$, for $w^J \in \mathcal{K}^J$ we define the following vector in $\RR^{m_J}$ as the finest-level \emph{nodal complementarity residual} which is defined as follows in the four cases (inactive, lower active, upper active, boundary):\footnote{Formally-speaking, $\hat r:\mathcal{V}^J \to \RR^{m_J}$ is simply a computable nonlinear map derived from $f^J$ and the vectors $\ell^J$, $\underline{\gamma}^J$, and $\overline{\gamma}^J$.  It is an important part of the implementation.}
\begin{equation}
\hat r(w^J)_p = \begin{cases}
    \ip{f^J(w^J)-\ell^J}{\psi_p}, & \underline{\gamma}^J(x_p) < w^J(x_p) < \overline{\gamma}^J(x_p), \\
    \min\{\ip{f^J(w^J)-\ell^J}{\psi_p},0\}, & w^J(x_p) = \underline{\gamma}^J(x_p), \\
    \max\{\ip{f^J(w^J)-\ell^J}{\psi_p},0\}, & w^J(x_p) = \overline{\gamma}^J(x_p), \\
    0, & x_p \in \partial_D\Omega. \end{cases} \label{eq:cpresidual}
\end{equation}
The point is that $\hat r(w^J)\in \RR^{m_J}$ is nonzero if there are nodes where MCP \eqref{eq:fe:mcp} is violated, either because the pointwise classical residual $\ip{f^J(w^J)-\ell^J}{\psi_p}$ remains nonzero where the constraints are inactive, or because it has the wrong sign on an actively-constrained degree of freedom.

A solver for \eqref{eq:fe:vi} using repeated applications of \pr{nmcd-vcycle} would start with an initial iterate $(w^J)^{(0)}$, do V-cycles, and stop once the finest-level nodal residual $\hat r(w^J)$ was sufficiently small.  For example, using tolerances \id{atol}$>0$ and \id{rtol}$>0$ the stopping criterion might be
\begin{equation}
\|\hat r(w^J)\| < \id{atol} \qquad \text{or} \qquad \frac{\|\hat r(w^J)\|}{\|\hat r((w^J)^{(0)})\|} < \id{rtol},
\end{equation}
for some norm $\|\cdot\|$ on $\RR^{m_J}$.  A step tolerance could also be used, measuring the difference between successive iterates.

@@

Algorithm \ref{alg:nmcd} in the main text, \pr{nmcd-vcycle}, generalizes two better-known multilevel algorithms.  First we show that it reduces to the FAS multigrid V-cycle for nonlinear PDEs (which itself generalizes the standard linear multigrid V-cycle \cite{Trottenbergetal2001}).  Second we reduce NMCD to an existing MCD V-cycle for unilateral linear variational inequalities \cite{GraeserKornhuber2009}.  This Appendix helps readers understand Algorithm \ref{alg:nmcd} by separating those assumptions which relate to nonlinear and FAS aspects from the nontrivial modifications of standard multigrid which are caused by inequality constraints.  The later arise even when $f$ in \eqref{eq:vi} is a linear operator.

To derive the classical FAS multigrid V-cycle from Algorithm \ref{alg:nmcd}, we remove the inequality constraints from VI problem \eqref{eq:fe:vi} and then simplify the pseudocode accordingly.  To do this we suppose that the finest-level obstacles $\underline{\gamma}^J$ and $\overline{\gamma}^J$ are very negative and positive, respectively, with magnitude exceeding $u^J$ or any prospective iterate $w^J$.  Finest-level problem \eqref{eq:fe:vi} is then unconstrained except for its Dirichlet boundary conditions.  Let $\mathcal{V}_0^J \subset \mathcal{V}^J$ denote the space with zero values on $\partial_D\Omega$.  By replacing ``$v-u^J$'' with an arbitrary $v \in \mathcal{V}_0^J$ we have the finest-level, weak form problem
\begin{equation}
\ip{f^J(u^J)}{v} = \ip{\ell^J}{v} \qquad \text{for all } v\in \mathcal{V}_0^J \label{eq:app:fas:pde}
\end{equation}
corresponding to the strong-form PDE $f(u)=\ell$.

Furthermore, the downward $\mathcal{D}^j$ and upward $\mathcal{U}^j$ constraint sets can be arbitrarily enlarged so that the coarser-level corrections $y^j$ and $z^j$ in Algorithm \ref{alg:nmcd} are also never constrained.  Formally-speaking we might do this, still satisfying the LDC ordering property \eqref{eq:fe:chiordering}, by choosing LDCs with all gaps being large so that $\underline{\chi}^j \ll 0 \ll \overline{\chi}^j$ and also $\underline{\phi}^j = \underline{\chi}^j - \underline{\chi}^{j-1} \ll 0 \ll \overline{\phi}^j = \overline{\chi}^j - \overline{\chi}^{j-1}$.\footnote{As noted after \eqref{eq:fe:chiordering}, the application of monotone restriction operators $\maxR,\minR$ is not required to generate LDCs $\underline{\chi}^j$ and $\overline{\chi}^j$ satisfying the ordering property.}

Now the correction VIs \eqref{eq:fe:downvi} and \eqref{eq:fe:upvi} are also unconstrained, and lines 2--3 and 6--9 can be removed from Algorithm \ref{alg:nmcd}.  Also, because the downward corrections $y^j$ are no longer separately constrained, we may simply refer to the corrected solution $w^j=g^j+y^j$ instead of $g^j$ and $y^j$ separately.  Then the smoothers and the coarsest-level solver can be regarded as in-place modifications of $w^j$.  Now noting $g^{j-1}=\iR w^j$ in line 12, after down-smoothing, the coarsened source functional can be rewritten using notation $w^{j-1}=\iR w^j$:
\begin{equation}
\ell^{j-1} = f^{j-1}\left(w^{j-1}\right) + R\left(\ell^j-f^j(w^j)\right). \label{eq:app:fas:levelsource}
\end{equation}
(Compare line 13 of Algorithm \ref{alg:nmcd}.)

>>>>

In summary, inclusions \eqref{eq:fe:downwardsuminclusion} and \eqref{eq:fe:upwardsuminclusion} suffice to show the admissibility of all iterates and corrections in V-cycle Algorithm \ref{alg:fascd}.  Lemma \ref{lem:downwardadmissibility} explains downward admissibility, and Lemma \ref{lem:upwardadmissibility} upward admissibility, in Algorithm \ref{alg:fascd}.  However, a convergence analysis following \cite{Tai2003} or \cite{GraeserKornhuber2009}, somehow extending their arguments to VIs \eqref{eq:vi} which are not of minimization type, and/or to box constraints, would seem to require full CDs, and perhaps new ideas.

@@

To state \fascd Algorithm \ref{alg:fascd} below we must clarify the problems solved at each mesh level.  As in FAS multigrid for nonlinear PDEs \cite{BrandtLivne2011,Bruneetal2015,Trottenbergetal2001}, both the solution approximation and the residual must be coarsened, and a source functional created.  On mesh level $j$ we assume a re-discretized nonlinear operator $f^j$, with output in $(\cV^j)'$, which is an FE discretization of $f$ in \eqref{eq:boxdirichletvi}, presumably using quadrature and etc.~in the same scheme which generated $f^J$ in \eqref{eq:fe:vi}.  Note that finest-level problem \eqref{eq:fe:vi} uses the original source functional $\ell^J$, and an admissible iterate $w^J \in \mathcal{K}^J$ has already\footnote{Conceptually-speaking this is true.  In practice the LDCs $\underline{\chi}^j,\overline{\chi}^j,\underline{\phi}^j,\overline{\phi}^j$ are generated during the descent stage; see Algorithm \ref{alg:fascd}.  Observe that the LDCs do not depend on $w^j$ for $j<J$, nor on the corrections $y^j,z^j$.} been used to general the LDCs and constraint sets $\mathcal{D}^j$, $\mathcal{U}^j$ at each level (Section \ref{sec:ldcs}).

@@

If we were to inject the original box constraints \eqref{eq:fe:fineconstraintset} downward to coarser meshes, i.e.~$\underline{\gamma}^j = \iR \underline{\gamma}^{j+1}$ and $\overline{\gamma}^j = \iR \overline{\gamma}^{j+1}$, then, because injection preserves inequalities, and because $y^{j+1} \in \mathcal{D}^{j+1}$, it would follow from \eqref{eq:fe:definew} that $\underline{\gamma}^j \le w^j \le \overline{\gamma}^j$.  In this sense $w^j$ is admissible with respect to the box constraints, but, following \cite{GraeserKornhuber2009}, our use of LDCs (Table \ref{tab:ldcs}) avoids the need to generate such box constraints $\underline{\gamma}^j$ and $\overline{\gamma}^j$ on each level.  (Note that while $w^j \in \cV^j \subset \cV^J$ holds, generally $w^j \notin \cK^J$.)  In any case, Algorithm \ref{alg:fascd} will yield a final, finest-level iterate $w^J+z^J \in \cK^J$ which is genuinely admissible for problem \eqref{eq:fe:vi}.

Regarding formula \eqref{eq:fe:levelsource}, the formula for source functional $\ell^j$ is a classical FAS construction.  (Compare equation (8.5b) in \cite{BrandtLivne2011} or equation (5.3.14) in \cite{Trottenbergetal2001}.)  It has the following explanation, mimicking the PDE version, which simultaneously justifies \eqref{eq:fe:levelsource} and VI problems \eqref{eq:fe:downvi} and \eqref{eq:fe:upvi}.  First, the $j+1$ level problem \eqref{eq:fe:downvi} is not solved exactly by the finer-level correction $y^{j+1}$---the down-smoother was only approximate---so we seek a further correction $y$ on the $j+1$ level so that
\begin{equation}
\ip{f^{j+1}(w^{j+1}+y^{j+1}+y)}{v-y^{j+1}-y} \ge \ip{\ell^{j+1}}{v-y^{j+1}-y} \quad \text{\emph{(notional)}} \label{eq:fe:downvinotional}
\end{equation}
for all $v\in \mathcal{D}^j$.  However, $y$ will actually be approximated from solving a coarser, $j$th-level problem generated by modifying \eqref{eq:fe:downvinotional} in 4 steps:
\begin{enumerate}
\item subtract the computable quantity $f^{j+1}(w^{j+1}+y^{j+1}) \in (\mathcal{V}^{j+1})'$ from both sides,
\item replace the residual $\ell^{j+1}-f^{j+1}(w^{j+1}+y^{j+1})$ on the right by its restriction,
\item replace $w^{j+1}+y^{j+1}$, where it appears on the left, by its restriction,
\item and, where it appears on the left, replace $f^{j+1}$ by the coarser rediscretization $f^j$.
\end{enumerate}
Steps \emph{(ii)}--\emph{(iv)} are justified when the residual and the error are smooth as a result of the progress made by the $j+1$ level down-smoother \cite[subsection 5.3.4, for example]{Trottenbergetal2001}.  These steps now yield a problem for the $j$th-level correction $y^j \in \mathcal{D}^j$:
\begin{align}
&\ip{f^j\left(\iR(w^{j+1}+y^{j+1})+y^j\right)}{v-y^j} - \ip{f^j\left(\iR(w^{j+1}+y^{j+1})\right)}{v-y^j} \label{eq:fe:downvicluttered} \\
&\qquad \ge \ip{R\left(\ell^{j+1}-f^{j+1}(w^{j+1}+y^{j+1})\right)}{v-y^j} \notag
\end{align}
This is the same as VI problem \eqref{eq:fe:downvi}, but in cluttered form; definitions \eqref{eq:fe:definew} and \eqref{eq:fe:levelsource} yields \eqref{eq:fe:downvi}.  The explanation for VI \eqref{eq:fe:upvi} is essentially identical.

We will regard the smoothers for VI problems \eqref{eq:fe:downvi} and \eqref{eq:fe:upvi} as approximate, in-place procedures, acting on the corrections $y^j$ and $z^j$, respectively (Section \ref{sec:smoothers}).  Also, as is standard in multigrid, at the coarsest $j=0$ level we suppose that problem \eqref{eq:fe:upvi} is solved quite accurately.  However, this coarsest-level solver may be the same as the other smoothers, perhaps with a stronger tolerance.

These ideas come together in Algorithm \ref{alg:fascd}.  \pr{fascd-vcycle} acts in-place on $w^J$, and it assumes certain semantics for the \pr{smooth} and the coarsest-level \pr{solve} procedures (Section \ref{sec:smoothers}).  These do in-place modifications of their final arguments, without modifying their first four arguments.  $\pr{smooth}^{\id{down}}$ and $\pr{smooth}^{\id{up}}$ denote \id{down} and \id{up} iterations, respectively.

\begin{pseudofloat}[ht]
\begin{pseudo}
\pr{fascd-vcycle}(\ell^J,\underline{\gamma}^J,\overline{\gamma}^J;w^J)\text{:} \\+
    $\underline{\chi}^J = \underline{\gamma}^J - w^J {\large \strut}$ \\
    $\overline{\chi}^J = \overline{\gamma}^J - w^J {\large \strut}$ \\
    for $j=J$ downto $j=1$ \\+
      $\underline{\chi}^{j-1} = \maxR \underline{\chi}^j {\large \strut}$ \\
      $\overline{\chi}^{j-1} = \minR \overline{\chi}^j {\large \strut}$ \\
      $\underline{\phi}^j = \underline{\chi}^j - P\underline{\chi}^{j-1} {\large \strut}$ \\
      $\overline{\phi}^j = \overline{\chi}^j - P\overline{\chi}^{j-1} {\large \strut}$ \\
      $y^j = 0$ \\
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j,\underline{\phi}^j,\overline{\phi}^j,w^j;y^j)$  \ct{smoothing of \eqref{eq:fe:downvi} in $\mathcal{D}^j$}\\
      $w^{j-1} = \iR(w^j + y^j)$ \\
      $\ell^{j-1} = f^{j-1}(w^{j-1}) + R \left(\ell^j - f^j(w^j+y^j)\right)$ \\-
    $z^0 = 0$ \\
    $\text{\pr{solve}}(\ell^0,\underline{\chi}^0,\overline{\chi}^0,w^0;z^0)$ \hspace{1.0cm} \ct{solving of \eqref{eq:fe:upvi} in $\mathcal{U}^0$} \\
    for $j=1$ to $j=J$ \\+
      $z^j = y^{j} + P z^{j-1}$ \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j,\underline{\chi}^j,\overline{\chi}^j,w^j;z^j)$  \ct{smoothing of \eqref{eq:fe:upvi} in $\mathcal{U}^j$} \\-
    $w^J \gets w^J+z^J$
\end{pseudo}
\caption{The \fascd V-cycle as an iteration for solving FE VI problem \eqref{eq:fe:vi}.  $f^j$ denotes an FE discretization of $f$ in problem \eqref{eq:boxdirichletvi}.}
\label{alg:fascd}
\end{pseudofloat}

Regarding storage required for \pr{fascd-vcycle}, and assuming all box constraints are nontrivial, 7 vectors must be allocated on each level: $\underline{\chi}^j$, $\overline{\chi}^j$, $\underline{\phi}^j$, $\overline{\phi}^j$, $w^j$, $\ell^j$, $y^j$.  Note $z^j$ can use the same storage as $y^j$.  On the finest level one must also store $\underline{\gamma}^J$ and $\overline{\gamma}^J$, while on the coarsest level there are only 5 vectors because $\underline{\phi}^0=\underline{\chi}^0$ and $\overline{\phi}^0=\overline{\chi}^0$.  Therefore, using standard estimates for 2D refinement plus a geometric series argument \cite{Trottenbergetal2001}, the total floating-point storage, excluding the implementations of smoothers and the coarse-level solver, is
\begin{equation}
9 m_J + 7 m_{J-1} + \dots + 7 m_1 + 5 m_0 \approx \left(9 + 7 \sum_{k=1}^\infty \frac{1}{4^k}\right) m_J \approx 11 m_J.
\end{equation}
For comparison, a single-level method needs $4 m_J$ storage for the vectors $\underline{\gamma}^J,\overline{\gamma}^J,\ell^J,w^J$.

If some of the obstacles are trivial, for instance if $\overline{\gamma}^J=+\infty$ in a unilateral obstacle problem, storage is correspondingly reduced at all levels.  Furthermore, we will see in Section \ref{sec:results} that, though \pr{fascd-vcycle} defines a general $V(\pr{down},\pr{up})$ cycle for VI problems, in fact the $V(0,\pr{up})$ case generally produces the best performance.  In that case the storage is reduced to about $9 m_J$ by neglecting the downward LDCs $\underline{\phi}^j$, $\overline{\phi}^j$ at all levels.

Repeated application of \pr{fascd-vcycle} to solve VI \eqref{eq:fe:vi} will not reduce the finest-level residual $f^J(w^J) - \ell^J$ to zero everywhere because the problem is inequality-constrained.  We need to identify the vector quantity whose norm must be controlled by a tolerance.  Recalling mixed complementarity problem \eqref{eq:fe:mcp}, and denoting the hat function at node $x_p \in \mathcal{T}^J$ by $\psi_p$, for $w^J \in \mathcal{K}^J$ we define the finest-level \emph{nodal complementarity residual} in $\RR^{m_J}$.  It is defined in the four cases listed after \ref{eq:fe:mcp}, namely inactive, lower active, upper active, and boundary:
\begin{equation}
\hat r(w^J)_p = \begin{cases}
    \ip{f^J(w^J)-\ell^J}{\psi_p}, & \underline{\gamma}^J(x_p) < w^J(x_p) < \overline{\gamma}^J(x_p), \\
    \min\{\ip{f^J(w^J)-\ell^J}{\psi_p},0\}, & w^J(x_p) = \underline{\gamma}^J(x_p), \\
    \max\{\ip{f^J(w^J)-\ell^J}{\psi_p},0\}, & w^J(x_p) = \overline{\gamma}^J(x_p), \\
    w^J(x_p) - g_D^J(x_p), & x_p \in \partial_D\Omega. \end{cases} \label{eq:cpresidual}
\end{equation}
Where the box constraints pinch at $x_p \notin \partial_D\Omega$, i.e.~where $\underline{\gamma}^J(x_p)=u^J(x_p)=\overline{\gamma}^J(x_p)$, we define $\hat r(w^J)_p=0$.  Note that $\hat r:\mathcal{V}^J \to \RR^{m_J}$ is a computable nonlinear map derived from $f^J$ and the vectors $\ell^J$, $\underline{\gamma}^J$, and $\overline{\gamma}^J$.  For unconstrained PDEs one would use only the inactive and boundary cases.

The idea here is that $\hat r(w^J)$ is nonzero only if there are nodes where MCP \eqref{eq:fe:mcp} is violated, either because the pointwise residual $\ip{f^J(w^J)-\ell^J}{\psi_p}$ remains nonzero where the constraints are inactive, or because it has the wrong sign on an actively-constrained degree of freedom, or because the Dirichlet boundary condition is not satisfied.

A solver for \eqref{eq:fe:vi} using repeated applications of \pr{fascd-vcycle} would start with an initial iterate $w_{(0)}^J$, do V-cycles, and stop once the finest-level nodal residual $\hat r(w^J)$ was sufficiently small.  For example, using tolerances \id{atol}$>0$ and \id{rtol}$>0$ the stopping criterion might be
\begin{equation}
\|\hat r(w^J)\| < \id{atol} \qquad \text{or} \qquad \frac{\|\hat r(w^J)\|}{\|\hat r((w^J)^{(0)})\|} < \id{rtol},
\end{equation}
for some norm $\|\cdot\|$ on $\RR^{m_J}$.  A step tolerance could also be used, measuring the difference between successive iterates.

@@

Algorithm \ref{alg:fascd} in the main text, \pr{nmcd-vcycle}, generalizes two better-known multilevel algorithms.  First we show that it reduces to the FAS multigrid V-cycle for nonlinear PDEs (which itself generalizes the standard linear multigrid V-cycle \cite{Trottenbergetal2001}).  Second we reduce NMCD to an existing MCD V-cycle for unilateral linear variational inequalities \cite{GraeserKornhuber2009}.  This Appendix helps readers understand Algorithm \ref{alg:fascd} by separating those assumptions which relate to nonlinear and FAS aspects from the nontrivial modifications of standard multigrid which are caused by inequality constraints.  The later arise even when $f$ in \eqref{eq:vi} is a linear operator.

To derive the classical FAS multigrid V-cycle from Algorithm \ref{alg:fascd}, we remove the inequality constraints from VI problem \eqref{eq:fe:vi} and then simplify the pseudocode accordingly.  To do this we suppose that the finest-level obstacles $\underline{\gamma}^J$ and $\overline{\gamma}^J$ are very negative and positive, respectively, with magnitude exceeding $u^J$ or any prospective iterate $w^J$.  Finest-level problem \eqref{eq:fe:vi} is then unconstrained except for its Dirichlet boundary conditions.  Let $\mathcal{V}_0^J \subset \mathcal{V}^J$ denote the space with zero values on $\partial_D\Omega$.  By replacing ``$v-u^J$'' with an arbitrary $v \in \mathcal{V}_0^J$ we have the finest-level, weak form problem
\begin{equation}
\ip{f^J(u^J)}{v} = \ip{\ell^J}{v} \qquad \text{for all } v\in \mathcal{V}_0^J \label{eq:app:fas:pde}
\end{equation}
corresponding to the strong-form PDE $f(u)=\ell$.

Furthermore, the downward $\mathcal{D}^j$ and upward $\mathcal{U}^j$ constraint sets can be arbitrarily enlarged so that the coarser-level corrections $y^j$ and $z^j$ in Algorithm \ref{alg:fascd} are also never constrained.  Formally-speaking we might do this, still satisfying the LDC ordering property \eqref{eq:fe:chiordering}, by choosing LDCs with all gaps being large so that $\underline{\chi}^j \ll 0 \ll \overline{\chi}^j$ and also $\underline{\phi}^j = \underline{\chi}^j - \underline{\chi}^{j-1} \ll 0 \ll \overline{\phi}^j = \overline{\chi}^j - \overline{\chi}^{j-1}$.\footnote{As noted after \eqref{eq:fe:chiordering}, the application of monotone restriction operators $\maxR,\minR$ is not required to generate LDCs $\underline{\chi}^j$ and $\overline{\chi}^j$ satisfying the ordering property.}

Now the correction VIs \eqref{eq:fe:downvi} and \eqref{eq:fe:upvi} are also unconstrained, and lines 2--3 and 6--9 can be removed from Algorithm \ref{alg:fascd}.  Also, because the downward corrections $y^j$ are no longer separately constrained, we may simply refer to the corrected solution $w^j=g^j+y^j$ instead of $g^j$ and $y^j$ separately.  Then the smoothers and the coarsest-level solver can be regarded as in-place modifications of $w^j$.  Now noting $g^{j-1}=\iR w^j$ in line 12, after down-smoothing, the coarsened source functional can be rewritten using notation $w^{j-1}=\iR w^j$:
\begin{equation}
\ell^{j-1} = f^{j-1}\left(w^{j-1}\right) + R\left(\ell^j-f^j(w^j)\right). \label{eq:app:fas:levelsource}
\end{equation}
(Compare line 13 of Algorithm \ref{alg:fascd}.)

At the completion of down-smoothing the solution iterate on the $j$th level is $w^j=g^j+y^j$.  Now we must be careful that going upward only the coarse correction, and not the coarse solution, gets prolonged.\footnote{Compare Remark 5.3.9 in \cite{Trottenbergetal2001}, a warning to novices.}  The coarser-level iterate is $w^{j-1} = g^{j-1}+z^{j-1}$ after up-smoothing on the $j-1$ level.  From Algorithm \ref{alg:fascd} the initial value of the correction $z^j$, before smoothing, is $z^j = y^j + P z^{j-1}$.  These facts imply that the $j$th-level iterate before up-smoothing must be set to
    $$w^j \gets w^j + Pz^j = (g^j+y^j)+P(w^{j-1} - g^{j-1}) = w^j + P(w^{j-1} - \iR w^j).$$
This is a standard FAS statement.  In fact, with the given modifications, we have the much-simplified pseudocode in Algorithm \ref{alg:fas}.  This FAS V-cycle algorithm appears in various places, for example Algorithm 14 of \cite{Bruneetal2015} or in section 5.3.4 of \cite{Trottenbergetal2001}.
